---
title: "Weeks 11 & 12 Exercises"
author: "Stewart Wilson"
date: "`r Sys.Date()`"
output: pdf_document
---
# Introduction to Machine Learning
```{r loading-packages, echo=FALSE, include=FALSE}
library(ggplot2)
library(caTools)
library(class)
library(caret)
```
## Scatter Plot of Data
```{r echo=FALSE}
binary <- read.csv("C:/Users/Stewart/Documents/GitHub/dsc520/data/binary-classifier-data.csv")
trinary <- read.csv("C:/Users/Stewart/Documents/GitHub/dsc520/data/trinary-classifier-data.csv")
binary_plot <- ggplot(binary, aes(x=x, y=y, colour=label)) + geom_point()
trinary_plot <- ggplot(trinary, aes(x=x, y=y, colour=label)) + geom_point()
binary_plot
trinary_plot
```
## Training Binary Data Sets
```{r}
# split binary into test and training set
train_index <- createDataPartition(binary$label, p=.6)$Resample1
training_binary <- binary[train_index, ]
test_binary <- binary[-train_index, ]
# matrix for num of clusters
kmatrix <- c(3, 5, 10, 15, 20, 25)
# will collect accuracy count for clusters
accBin <- c()
# runs nearest neighbor for every given k in kmatrix
# calculates accuracy for each loops
index <- 0
for(i in kmatrix){
  index = i + 1
  kModBin <- knn(train=training_binary, test=test_binary, cl=training_binary$label, k=i)
  accBin[index] <- 100*sum(test_binary$label == kModBin)/NROW(test_binary$label)
}
# plotting accuracy vs k
accuracy_df <- data.frame(accBin)
accuracy_df$ID <- seq.int(nrow(accuracy_df))
accuracy_plot <- ggplot(data = accuracy_df, aes(x = ID, y = accBin)) + geom_point()
accuracy_plot
```
## Training Trinary Data Sets
```{r}
# split binary into test and training set
train_index <- createDataPartition(trinary$label, p=.6)$Resample1
training_trinary <- trinary[train_index, ]
test_trinary <- trinary[-train_index, ]
# matrix for num of clusters
kmatrix <- c(3, 5, 10, 15, 20, 25)
# will collect accuracy count for clusters
acc_Trin <- c()
# runs nearest neighbor for every given k in kmatrix
# calculates accuracy for each loops
index <- 0
for(i in kmatrix){
  index = i + 1
  kModBin <- knn(train=training_trinary, test=test_trinary, cl=training_trinary$label, k=i)
  acc_Trin[index] <- 100*sum(test_trinary$label == kModBin)/NROW(test_trinary$label)
}
# plotting accuracy vs k
accuracy_df2 <- data.frame(acc_Trin)
accuracy_df2$ID <- seq.int(nrow(accuracy_df2))
accuracy_plot2 <- ggplot(data = accuracy_df2, aes(x = ID, y = acc_Trin)) + geom_point()
accuracy_plot2

```
### Linear Model?
Looking back at the original graphs of the data, I do not think a linear classifier would work
well on these datasets. There is no clear straight line that could split either
dataset nicely so the model would most likely not be very accurate.

Looking back at the accuracy from last week's exercises (58%), it is clear
that clustering vastly improved the accuracy of the model.

# Clustering

