---
title: "What Effect Does the War on Terror Have On Terrorism?"
subtitle: "Final Project"
author: "Stewart Wilson"
date: "`r Sys.Date()`"
output: pdf_document
---
# I. Introduction
  Following the unprecedented attack on the Twin Towers on 9/11, the USA declared a 
War on Terror. They passed laws such as the PATRIOT ACT, invaded two sovereign nations,
expanded the surveillance apparatus, and cracked down domestically and abroad on
any and all hints of terrorist action. 20 years later, however, and questions of its efficacy, cost, and morality loom large. Setting questions of morality aside momentarily, 
what effect has the War on Terror had on terrorism? The answer to this question is of severe import because it offers insight into how the US should continue to operate. If the tactics
implemented are both costly and ineffective, it can hardly be justified to continue.
Furthermore, this is a question pertaining to data science because, through analyzing
data related to and influenced by War on Terror policies, we can determine the 
statistical effect that these policies have had on terrorism.

  The following exploratory data analysis is meant to be a beginning towards answering
some of these questions. It is by no means an exhaustive analysis. Indeed, as 
will be seen, as the data is explored, more questions are raised than are answered.
Yet, while there are no definitive answers given in this analysis, several insights
can be gained that could serve as signposts signalling the way for future data 
analysis.

  The structure of this paper follows the pathway of my thinking through the problem
of quantifying the effect the War on Terror has had on global terrorism rates.
I will lay out the problem in more details, summarize how I addressed it, explain
the insights I gained, the implications these insights raise, and the limitations
inherent to my analysis. I hope the reader will find this analysis as interesting
and impactful as I did. 

# II. The Problem at Hand (Step 1)
  At first, I began by posing the problem as broadly as possible: did the War on
Terror have a significant effect on terrorism worldwide and abroad. I immediately
ran into difficulties trying to break that question into manageable parts. What exactly
is the War on Terror? Can it be reduced to quantitative terms? How do you compare
terrorism globally to terrorism domestically? How do you compare domestic responses
to global ones?

  In the first step of my research, I decided to split the problem in two.
I would investigate domestic terror in relation to domestic responses and
I would investigate global terror in relation to U.S. global response. Domestically,
I would focus on the actions of Department of Homeland Security and globally on 
US military action.

  However, when I began the second step of my analysis, I had to further restrict the problem.
There just simply was not enough publicly available data to sufficiently investigate
the domestic response to terrorism. Moreover, I could only find data related to 
troop deployment in the European Union, which while helpful leaves out West Asia
where much of US troops were focused during the War on Terror.

  As such, I limited my analysis to investigating the relationship between 
US Defense Budget and global terrorism rates, with the Defense Expenses being a proxy
for the actions of the War on Terror. This problem was much more manageable for
the scope of the project, though it came with many limitations, as I will expand on
in Section VI.

# III. Addressing the Problem
  During the second step of my project, I imported data, cleaned it, and analyzed it.
  
## Data Importing and Cleaning (Step 2)
  I used 3 datasets during my analysis. The first was the [Global Terrorism Database](https://www.kaggle.com/datasets/START-UMD/gtd), which contained over 100
variables for each attack specifying things such as tactics, perpetrators, targets, and outcomes. I determined that the following variables were most relevant to my analysis:

- iyear: year attack conducted
- imonth: month attack conducted
- country_txt: country where attack took place
- region: region where attack took place
- success: whether the attack was successful
- nkill: number killed during attack

  As said in the pervious section, the Troop Deployment dataset proved less useful than
anticipated, so I left it aside to streamline my research.


  The third dataset contained US Defense Spending between 1960 and 2020.
I merged this dataset with Global Terrorism Database,
made the variable names uniform, and limited the years between 1970 and 2019.

The final dataset can be seen below.
```{r data-cleaning, echo=FALSE}
library(dplyr)
# import data
terrorism_db <- read.csv("C:/Users/Stewart/Documents/GitHub/dsc520/final_project/data/globalterrorismdb.csv")
military_spending <- read.csv("C:/Users/Stewart/Documents/GitHub/dsc520/final_project/data/MilitarySpending.csv")
terrorism2 <- terrorism_db %>% select(iyear, imonth, country_txt, region, success, nkill)
terror_war <- terrorism2 %>% full_join(military_spending, by=c('iyear'='Year'))
terror_war <- terror_war %>% rename_with(tolower)
terror_war <- terror_war %>% filter(iyear>=1970)
terror_war %>% 
  arrange(iyear) %>% 
  slice(c(1:5),(50000:50005),(175000:175005))
```

## Data Analysis (Step 2)
  Now that the dataset had been imported and cleaned, I could begin my analysis.
The plan was first to slice the data into attacks before 9/11 and after 9/11. Then,
observe comparative statistical characteristics between the slices. Finally, 
perform correlation analysis on the Defense Spending and Number of Terrorist Attacks.

  One problem I ran into at this stage was meaningfully comparing variables to one another.
Given that the defense spending values were repeated multiple times (more terrorist attacks
per year than number of defense budgets naturally), the effect of spending was vastly
overstated. To correct for this, I created a new dataset that contained one column
for defense spending, and one for total attacks per year. This allowed me a greater
ability to see the general relationship between attacks and spending.

  I created the following tables/graphs to visualize the results of my analysis.
Please note that the red line in the first three graphs mark 9/11.

```{r analysis-so-far, echo=FALSE, include=FALSE}
library(pastecs)
# split into before and after
before <- terror_war %>% filter(iyear <= 2001)
after <- terror_war %>% filter(iyear > 2001)
# num attacks before and after 9/11
before_attacks <- before %>% group_by(iyear) %>% summarize(nrow(iyear))
# Let's look at the defense budget!
before %>% arrange(desc(nkill))
spending_before <- military_spending %>% filter(Year <=2001) %>%  summarize(sum(DefenseBudget), mean(DefenseBudget))
spending_after <- military_spending %>% filter(Year > 2001) %>%  summarize(sum(DefenseBudget), mean(DefenseBudget))
# us terror attacks
us_terrorism <- terror_war %>% filter(country_txt=="United States") 
us_terrorismb <- us_terrorism %>% filter(iyear <= 2001)
us_terrorisma <- us_terrorism %>% filter(iyear>2001)
# correlation tests!!
attack_yr <- terror_war %>% dplyr::count(iyear)
spending <- military_spending %>% filter(Year>=1970)
war_cor <- full_join(attack_yr, spending, by=c("iyear"="Year"))
cor1 <- cor.test(attack_yr$n, spending$DefenseBudget, method="spearman", conf.level = .99, exact = FALSE)
cor2 <- cor(attack_yr$n, spending$DefenseBudget, method="spearman")
r_sq <- cor2 ** 2
# find p-values
# p-value < .005 so not likely due to chance
```

```{r visualizations, echo=FALSE}
# attacks by year
library(ggplot2)
attacks_hist <- ggplot(data=terror_war)+ geom_histogram(aes(x=iyear), binwidth = 1, fill='orchid4') + xlab("Year") + ylab("Terrorist Attacks") + ggtitle("Terrorist Attacks by Year (1970-2019)") + geom_vline(xintercept = 2001, colour='red')
attacks_hist

# us attacks by year
attacks_us <- ggplot(data=us_terrorism) + geom_histogram(aes(x=iyear), binwidth=1, fill='blue') + xlab("Year") + ylab("Terrorist Attacks") + ggtitle("US Terrorist Attacks by Year (1970-2019)") + geom_vline(xintercept = 2001, colour='red')
attacks_us

# military expenditure bar
expenses <- ggplot(data = military_spending, aes(x=Year, y=DefenseBudget)) + geom_bar(stat="identity", fill='green4')+ xlab("Year") + ylab("Defense Budget") + ggtitle("US Military Expenditure by Year (1960-2020)") + geom_vline(xintercept=2001, colour='red')
expenses

# plotting military expenditure against terrorist attacks
scatter1 <- ggplot(war_cor, aes(x=DefenseBudget, y=n)) + geom_point(aes(color=iyear)) + xlab("Defense Budget") + ylab("Terrorist Attacks Per Year") +
  ggtitle("Relationship Between Defense Budget and Terrorist Attacks")
scatter1
```
  And here is a correlation table of GDP, US Population, # of Attacks, and Defense Budget
in order to place the correlation of defense budget and attacks into context.
```{r table, echo=FALSE}
cor_table <- cor(war_cor, method="spearman")
colnames(cor_table) <- c("Year", "Attacks", "DefenseBudget", "GDP", "Population")
rownames(cor_table) <- c("Year", "Attacks", "DefenseBudget", "GDP", "Population")
knitr::kable(cor_table)
```

### Regression Analysis
  Observing the scatterplot above, it appears the Defense Budget and Attacks Per Year
follow a linear relationship. This motivated me to do a simple regression analysis.
The below graph is the scatterplot above with a regression line fitted to it. One
can see that it is a fairly good fit, though there are some outliers in the 600 billion 
dollar spending domain.

```{r regression, echo=FALSE}
# simple regression outcome: Attacks per year, predictor: defense budget
terror_lm <-  lm(n ~ DefenseBudget, data=war_cor)
scatter2 <- ggplot(war_cor, aes(x=DefenseBudget, y=n)) + geom_point(aes(color=iyear)) +  geom_smooth(method="lm") + xlab("Defense Budget") + ylab("Terrorist Attacks Per Year") + ggtitle("Defense Budget vs Terrorist Attacks w/ Regression Line")
scatter2
```

## Recommendation for Models
  I would recommend doing time series analysis and building a time series model of 
terrorist attacks per year. This would help in identifying if there are  other
reasons for rising terrorist attacks outside of Defense Spending. For instance,
if particular years correlate with increased terrorist attacks the next year, one
could investigate events that may have occurred that explain that rise. In this way,
we can begin to build a more robust idea of the relationship between US military action
and terror.

  The model would be useful in trying to predict the amount of terrorist attacks in 
future years. This model could be compared to the regression model to see which
is more accurate. If the time series model is more accurate than the regression model,
it is again likely there is a third (or more) relation with changing terrorist rates.

# IV. Insights Gained
  There were a number of insights gained from the preceding analysis. Military 
expenditure skyrocketed post 9/11, which is suggestive of an increase in military
action in accordance with the War on Terror. Terrorist attacks too, increased
after 2001, though they decreased domestically in the US.

  The most striking insight gained by far from the analysis is the positive
correlation between US defense spending and terrorist attacks per year. This implies
that as defense spending rose, so too did the number of terrorist attacks. The analysis
also showed that this relation is not likely due to chance, meaning that the effect
of spending on terrorist attacks is statistically significant. The regression analysis
confirms this effect, with defense spending being an excellent predictor of terrorist attacks.

# V. Implications
  This last insight has remarkable implications for those wishing to understand
the effect of the War on Terror. If it is assumed that defense spending
is a good proxy for military action and that military action is the foremost
feature of the War on Terror, then the fact that increasing military spending 
has a positive relation to terrorist attacks seems to suggest that the War on Terror
is having the opposite effect! If such an account is true, then the implication
is that there needs to be some serious re-evaluation towards the means the US
has taken to try and curb terrorism.

  Of course, one could point out that correlation has no say in the direction of the relation.
It is equally plausible that the increased terrorist attacks cause more to be spent
on the military. This is possible, yet if it is true it does not mean therefore that the War on Terror
has been effective. That there has not been a consistent decrease in terrorist attacks
in the 20 years since 9/11 even when military spending has nearly quadrupled still
indicates that these methods are not working well. The question remains, how much more
money should continue to be put towards a war that has not made much progress towards
victory.

# VI. Limitations
  There are several limitations to my analysis that temper these implications. I
mentioned that we should assume that defense spending is a good proxy for military action
and therefor the War on Terror. This assumption may not hold. If the increase in
defense spending was spent more on research than on actual troop deployment, then
defense spending is not a perfect proxy for the War on Terror and thus the preceding
implications hold little water.

  In addition, given that the data spans the course of 50 years, the standards
for data collection probably shifted a large amount in that time. If data collection
was not consistent, then we cannot be sure that the data analysis was valid. There
remains the possibility that the reason terrorist attacks increased post 9/11 was
because data on those attacks were better collected in the the 2000s than the 1970s.

  If one wanted to improve this analysis, they would definitely need to gather
more data. If they could find datasets on troop deployment in West Asia, technology
changes as a result of the War on Terror, quantitative or qualitative impact of policies 
implemented post 9/11 then one would gain a more holistic knowledge of what constitutes
the War on Terror. One could do a similar analysis to what I did above but zooming
in on individual regions and seeing what impact the War on Terror had there. This 
would also aid in accounting for the data collection possibility,
since one would be able to just look at data after 2001.

  Another way to improve the analysis would be 

# VII. Concluding Remarks
  In spite of these limitations—or perhaps because of them—the analsyis conducted
above provided new insights into future research possibilities. It offers the possibility
of determining the statistical effects of the past 20 years of the War on Terror.
It opens up new avenues for understanding the relationship between U.S. foreign
policy and acts of terror. 

  On a final note, one should take care not to become too enraptured by the data
to forget that what is at stake here is not just the answer to a statistical question.
The choices that are made when it comes to the War on Terror and addressing terrorism have tangible effects.
These effects are felt in lives lost, buildings destroyed, and communities upended.
A true measure of the effects of the War on Terror should not just consider whether
the War on Terror accounts for a sufficient amount of variance in global terrorist rates,
but also whether that effect outweighs the secondary effects it has on the places
and people it turns its attention towards.